Connectome inferences using partial correlation from imaging calcium signal

### 2014-06-13 article.tex
rewrote section on partial correlation

Finally since there is a high correlation between neurons, we made a
low-rank hypothesis and perform noise filtering through a principal component
analysis. From the $p$ eigen vectors, we only kept $80\%$ of
the eigen vectors which have the largest eigen value. In the context of the
challenge, this assumption was deduced from ground truth of provided network.

\subsection{Network inference through partial correlation}
\label{sec:inference}

% The main challenge in the inference of the network $y$
% is to distinguish direct from indirect interactions \cite{de2004discovery}.
% For instance with the simple interaction graph $A \rightarrow B \rightarrow C$,
% we want to distinguish direct interactions $A \rightarrow B$ and $B \rightarrow C$
% from the indirect interaction $A \rightarrow C$.

From a probabilistic point of view, the inference of the undirected network
can be formulated as inferring conditional dependences and independences. Even
if \textit{correlation does no imply causation}[todo cite], it gives insight
on the underlying connectivity network.

Let us consider the neurons as random variables $\mathcal{X}_i$ which follows
a probability density function $p_{\mathcal{X}_i}$ and let us denote by
$p_{\mathcal{X}_i,\mathcal{X}_j|\mathcal{Z}}$ the joint probability
distribution between neurons $i$ and $j$ conditioned on a subset of neurons
$\mathcal{Z}$ (possibly empty). Two neurons $X_i$, $X_j$
will be said conditionally independent to a set of neurons
$\mathcal{Z}$ (possibly empty), denoted by  $X_i \perp X_j | \mathcal{Z}$,  if
$p_{\mathcal{X}_i,\mathcal{X}_j|\mathcal{Z}} = p_{\mathcal{X}_i|\mathcal{Z}}
p_{\mathcal{X}_j|\mathcal{Z}}$. The reciprocal is not true without
additional hypotheses.  Let us define $\mathcal{V}$ the set of all
neurons $\mathcal{X}_1, \dots, \mathcal{X}_p$, and by $\mathcal{V}^{-i,j}$ the
same set without neurons $\mathcal{X}_i$ and $\mathcal{X}_j$.

The connectivity network to infer is only made of direct relationships
which defined relation between neurons that still exists when
conditioned on all other neurons. In other words, there is not intermediary
between those two neurons. Under the hypothesis that neurons follow a jointly
Gaussian distribution $p_{\mathcal{X}_1, \dots, \mathcal{X}_p} \sim
\mathcal{N}(\mu; \Sigma)$, two neurons $\mathcal{X}_i$ and $\mathcal{X}_j$ are
independent $\mathcal{X}_i\perp \mathcal{X}_j$ if the Pearson correlation
coefficient $\rho_{X_iX_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii}
\Sigma_{jj}}} = 0,$ and are conditionally independant to all others neurons
$\mathcal{X}_i \perp \mathcal{X}_j |\mathcal{V}^{i,j}$ if the partial
correlation coefficient (of order $p-2$) $\rho_{X_i, X_j | \mathcal{V}^{i,j}}
= \frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0$,
where $\Sigma^{-1}$ is the inverse of the covariance matrix.
Therefore, by definition, the partial correlation coefficient between two
neurons only estimates the direct relationship between them.

In previous study \cite{shipley2002cause}, it has been suggested to
consider correlation coefficients of all possible orders,
i.e. conditioned on every subsets of the set of $p-2$ other variables. This
would take into account multiple indirect paths between a given pair of
variables. An obvious limitation to estimate correctly high-order coefficients
is the number of samples. However given the high number of samples $n$
in the Connectomics challenge, we only consider the highest order of partial
correlation coefficient.

\subsection{Stacking partial correlation matrix}
\label{sec:stacking}

% Speak about why averaging
The quality of the signal filtering and network inference is dependent upon
good hyper-parameter selection, which are themselves dependent on the properties
of the targeted network and measurement conditions, e.g. signal to noise ratio.
Ensemble methods are known to improve robustness with respect to parameter
selection [todo cite] and improve performance [todo cite].
A weighted average $\hat{y}$ of partial correlation matrices was estimated
from the timeseris $X \in \mathcal{R}^{n \times p}$
by varying unifthe hard-threshold value $\tau \in \mathcal{T}$  and the design
of the low-pass filter (coefficients
$(c_l)_{-3}^2 \forall l \in \left\{1,\ldots,4\right\}$)
\begin{align}
\hat{y} = \frac{1}{\sum_{l=1}^4 w_l |\mathcal{T}|} \sum_{l=1}^4  w_l \sum_{\tau \in \mathcal{T}} G^{\tau,m}(X)
\end{align}
where $G^{k, \tau}(X)$ gives the connectivity score matrix for
a given  $\tau$ and a set of coefficient $c_l$. following
the methodology described in previous section, associated to a weight
$w_p$ whose values are $w_1 = 0.01 , w_2 =1, w_3 = 0.7, w_4 = 0.9$.

@@

>>>>

Finally since there is a high correlation between neurons, we made a
low-rank hypothesis and perform noise filtering through a principal component
analysis. From the $p$ eigen vectors, we kept $80\%$ of
the eigen vectors which have the largest eigen value. In the context of the
challenge, this assumption was deduced from ground truth of provided network.

\subsection{Network inference through weighted average partial correlation
            coefficients}
\label{sec:inference}

Let us denote the fluorescence calcium indicators of all neuron
as a set of random variables $V = \left\{V_1, \ldots, V_{|V|}\right\}$, which follows
a joint probability distribution $p_\mathcal{V}$, and by
$V^{-i,j}$ the set of all variables except $V_i$ and $V_j$.
Two random variables $V_i$ and $V_j$ are said to be conditionally independent
to a (possibly empty) set of random variables $Z$, denoted by $V_i \perp V_j | Z$,
if the joint conditional probability distribution factorizes
$p_{\mathcal{V}_i, \mathcal{V}_j|\mathcal{Z}} = p_{\mathcal{V}_i|\mathcal{Z}}
p_{\mathcal{V}_j|\mathcal{Z}}$.  The inference of the undirected network
can be formulated as inferring conditional dependences
$V_i \not\perp V_j | V^{-i,j}$ between all pair of fluorescence signal
$i$ and $j$, in other words the inferred network will be
$\hat{A}: \hat{a}_{i,j} = I(V_i \not\perp V_j | V^{-i,j}), i, j \in V$.

Since all neurons are assumed to be connected to at least one neuron,
we thus have by definition $V_i \not\perp V_j$. However,  two random
variables $A$ and $B$ might be dependent taken alone ($A \not\perp B$), but not
once we observe a third variables $C$ ($A \perp B | C$).
The conditioning over all other random variables $V^{-i,j}$ is thus necessary
to avoid predicting an edge when two neurons are not directly connected.

If the random variables $V$ follows a joint Gaussian distribution
$p_\mathcal{V} \sim \mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and
covariance matrix $\Sigma$, we have the conditional independence
$V_i \perp V_j | V^{-i,j}$ if the partial correlation coefficient
\[
\rho_{V_i, V_j | V^{-i,j}}
= \frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}.
\]
is equal zero. The partial correlation gives a connectivity score
$\hat{A}: \hat{a}_{i,j} = \rho_{V_i, V_j | V^{-i,j}}, i, j \in V$ between all
pair of neuron $i$ and $j$. One could retrieve an estimated graph by thresholding
the matrix $\hat{A}$.
By contrast, we also have $V_i \perp V_j$ if the Pearson correlation
coefficient
\[
\rho_{V_i,V_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii}
\Sigma_{jj}}}
\]
is equal to zero.

% Speak about why averaging
The quality of the signal filtering and network inference depends
good hyper-parameter selection, which are themselves dependent on the properties
of the targeted network and measurement conditions, e.g. signal to noise ratio.
In order to improve robustness, a weighted average of partial correlation
matrices was estimated from the time series by varying the hard-threshold value
$\tau$ and the design of the low-pass filter (coefficients $(c_l^k)_{l=-3}^k$).

@@

In previous study \cite{shipley2002cause}, it has been suggested to
consider correlation coefficients of all possible orders,
i.e. conditioned on every subsets of the set of $p-2$ other variables. This
would take into account multiple indirect paths between a given pair of
variables.
